ExpressionView

### 2009-12-02 RExpressionView/inst/doc/ExpressionView.tutorial.Rnw
ExpressionView: Bugfix in Makefile. Minor changes.

\section{Export}
The \Rfunction{ExportEV} function allows the user to combine the available data and export it to an XML file that can be read by the Flash applet:
<<export, eval=FALSE>>=
ExportEV(modules, ALL, optimalorder, filename="file.evf")
@
The function gathers the data contained in the \Rclass{ExpressionSet} \variable{ALL}, orders it according to the optimal arrangement 
\variable{optimalorder} and adds the biclusters defined in \variable{modules}. The output is an uncompressed XML file that can be opened with any text viewer. We have chosen to use the extension \variable{.evf} (for ExpressionView file) for the data files. This extension is associated with the stand-alone version of the viewer, so that one can simply double-click on such a file to launch the program and load the data. The file association is the reason why we do not use the \variable{.xml} extension. A description of the XML layout can be found on the  href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}. Before exporting the data, the 
\Rfunction{ExportEV} function automatically calculates GO~\cite{ashburner00} and KEGG~\cite{kanehisa04} enrichments for the given biclusters.

@@

OpenEV()

@@

	\item[c] Exports the data of the currently viewed module(=bicluster) to a CSV file, that can be opened as a spreadsheet.

@@

	\item[e, f] Zoom and pan modes allow you to restrict the view to a particular part of the gene expression data. They behave like the functions used in graphic programs.

@@

	\item[k] Sets the visibility of the modules layer. Moving the slider to the left fades out the gene expression data, focusing on the Biclusters, while towards the opposite direction, the gene expression data moves to the foreground.

@@

>>>>

\section{Export}
The \Rfunction{ExportEV} function allows the user to combine the available data and export it to an XML file that can be read by the Flash applet:
<<export, eval=FALSE>>=
ExportEV(modules, ALL, optimalorder, filename="file.evf")
@
The function gathers the data contained in the \Rclass{ExpressionSet} \variable{ALL}, orders it according to the optimal arrangement 
\variable{optimalorder} and adds the biclusters defined in \variable{modules}. The output is an uncompressed XML file that can be opened with any text viewer. We have chosen to use the extension \variable{.evf} (for ExpressionView file) for the data files. This extension is associated with the stand-alone version of the viewer, so that one can simply double-click on such a file to launch the program and load the data. The file association is the reason why we do not use the \variable{.xml} extension. A description of the XML layout can be found on the \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}. Before exporting the data, the 
\Rfunction{ExportEV} function automatically calculates GO~\cite{ashburner00} and KEGG~\cite{kanehisa04} enrichments for the given biclusters.

@@

LaunchEV()

@@

	\item[c] Exports the data of the currently viewed module (=bicluster) to a CSV file, that can be opened as a spreadsheet.

@@

	\item[e, f] Zoom and pan modes allow you to restrict the view to a particular part of the gene expression data.

@@

	\item[k] Sets the visibility of the modules layer. Moving the slider to the left fades out the gene expression data, thus focusing on the Biclusters, while towards the opposite direction, the gene expression data moves to the foreground.

@@

\section{Using ExpressionView with non-gene expression data}
While ExpressionView is designed to work with gene expression data available in the form of a \Rpackage{Bioconductor} \Rclass{ExpressionSet}, it can also be used to visualize other data. Let us for instance use in-silico data generated by the \Rpackage{isa} package with dimensions 50 $\times$ 500 containing 10 overlapping modules:
<<in-silico>>=
library(ExpressionView)
# generate in-silico data with dimensions m x n containing M overlapping modules
# and add some noise
m <- 50
n <- 500
M <- 10
data <- isa.in.silico(num.rows=m, num.cols=n, num.fact=M, noise=0.1, overlap.row=5)[[1]]
modules <- isa(data)
@

The \Rfunction{ExportEV} uses the named list provided by the \variable{description} variable to label the data. First, let us annotate the rows and columns of the data set
<<data set annotation>>=
v <- vector("character"); for ( i in 1:dim(data)[1] ) { v[i] <- paste("row ", i, sep="") }; rownames(data) <- v
v <- vector("character"); for ( i in 1:dim(data)[2] ) { v[i] <- paste("column ", i, sep="") }; colnames(data) <- v
@

Next, we assign the meta data associated with the rows of the data matrix. In this example we use 5 tags labelled ''row tag``:
<<rowdata annotation>>=
rowdata <- matrix(nrow=dim(data)[1], ncol=5)
rownames(rowdata) <- rownames(data)
v <- vector("character"); for ( i in 1:dim(rowdata)[2] ) { v[i] <- paste("row tag ", i, sep="") }; colnames(rowdata) <- v
for ( i in 1:dim(rowdata)[1] ) { for ( j in 1:dim(rowdata)[2] ) { rowdata[i,j] <- paste("row description (", i, ", ", j, ")", sep="") }}
@
And similarly for the columns, using 10 "column tags``:
<<coldata annotation>>=
coldata <- matrix(nrow=dim(data)[2], ncol=10)
rownames(coldata) <- colnames(data)
v <- vector("character"); for ( i in 1:dim(coldata)[2] ) { v[i] <- paste("column tag ", i, sep="") }; colnames(coldata) <- v
for ( i in 1:dim(coldata)[1] ) { for ( j in 1:dim(coldata)[2] ) { coldata[i,j] <- paste("column description (", i, ", ", j, ")", sep="") }}
@

To finish  the description, we add some general information and merge it with the above tables to get a single named list:
<<in-silico data annotation>>=
description <- list(
experiment=list(
	title="Title", 
	xaxislabel="x-Axis Label",
	yaxislabel="y-Axis Label",
	name="Author", 
	lab="Address", 
	abstract="Abstract", 
	url="URL", 
	annotation="Annotation", 
	organism="Organism"),
coldata=coldata,
rowdata=rowdata
)
@
When dealing with gene expression data, the \variable{xaxislabel} is equal to ''genes`` and the  \variable{yaxislabel} is ''samples``. Finally, we export the data set to an ExpressionView file:
<<in-silico data export, eval=FALSE>>=
ExportEV(modules, data, filename="file.evf", description=description)
@

Simply load this file with the Flash applet and check where the various labels appear.
